---
title: "Two-Sample Means (Independent)"
subtitle: "STA5176 - Lecture 3"
author: "Dr. Seals"
format: revealjs
editor: visual
---

## Introduction

-   In the last lecture, we covered estimation and inference on a single mean using a one-sample *t*-test.

-   Today we will cover estimation and inference on two independent means using the two-sample *t*-test.

    -   We will cover two cases: equal variances and unequal variances.

-   We will only discuss the *t*-test approach, but note that under certain conditions, we could use a *z*-test.

## Data and Packages

```{r, warning = FALSE}
library(palmerpenguins)
library(tidyverse)

penguins <- palmerpenguins::penguins %>%
  mutate(sex = ifelse(sex == "male", "M", 
                      ifelse(sex == "female", "F", NA))) %>%
  filter(sex %in% c("M", "F"))

# I recoded male and female and removed missing data for display purposes
```

```{r, echo = TRUE, warning = FALSE}
library(palmerpenguins)
library(tidyverse)

head(penguins)
```

```{r, warning = FALSE}
penguins <- palmerpenguins::penguins

# reset the penguin dataset for examples
```

## Confidence Interval -- Definition

-   The CI for $\mu_1 - \mu_2$ assuming $\sigma_1^2 = \sigma_2^2$ is as follows:

$$ \left( \bar{x}_1-\bar{x}_2 \right) \pm t_{1-\alpha/2, \ n_1+n_2-2} s_p \sqrt{\frac{1}{n_1}+\frac{1}{n_2}}, $$

-   where $s_p$ is the pooled standard deviation and

$$ s_p=\sqrt{\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2}} $$

## Confidence Interval -- Definition

-   The CI for $\mu_1 - \mu_2$ assuming $\sigma_1^2 \ne \sigma_2^2$ is as follows:

$$ (\bar{x}_1-\bar{x}_2) \pm t_{1-\alpha/2, \text{df}} \sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}, $$

-   where df are *estimated* using Satterthwaite's approximation,

$$ \text{df}=\frac{ \left( \frac{s^2_1}{n_1} + \frac{s_2^2}{n_2} \right)^2 }{ \frac{(s_1^2/n_1)^2}{n_1-1} + \frac{(s_2^2/n_2)^2}{n_2-1} } $$

## Confidence Interval -- Comparing $s_i$

-   Generally, we do not need to worry about this, because R automatically assumes $\sigma^2_1 \ne \sigma^2_2$ and estimates the df.

    -   This is a safe approach because when we can assume $\sigma_1^2 = \sigma_2^2$, the estimated df is close to $n_1+n_2-2$.

-   For now, we can "eyeball" the standard deviations to determine equivalence.

    -   If one is more than twice the other, we will use Satterthwaite's approximation.

    -   If we want to use exact df, we can use the `var.equal` option in the [`t.test()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/t.test) function.

## Confidence Intervals -- R Syntax

-   We will use the [`t.test()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/t.test) function to find the CI,

```{r, echo = TRUE, eval = FALSE}
t.test([continuous variable] ~ [grouping variable],
       data = [dataset],
       conf.level = [confidence level])
```

## Confidence Intervals -- Penguins

-   Let's find a 95% CI on the average body mass for penguins, comparing females to males.

-   First, let's look at the standard deviations.

```{r, echo = TRUE}
penguins %>% 
  group_by(sex) %>%
  summarize(sd(body_mass_g, na.rm = TRUE))
```

-   Thus, we can assume that $\sigma_1^2 = \sigma_2^2$.

## Confidence Intervals -- Penguins

-   Now, to find the 95% CI for $\mu_{\text{F}} - \mu_{\text{M}}$,

```{r, echo = TRUE}
t.test(body_mass_g ~ sex,
       data = penguins,
       conf.level = 0.95,
       var.equal = TRUE)
```

-   The 95% CI is (-840.80, -526.02)

## Confidence Intervals -- Penguins

-   What happens if we do not specify $\sigma_1^2 = \sigma_2^2$?

```{r, echo = TRUE}
t.test(body_mass_g ~ sex,
       data = penguins,
       conf.level = 0.95)
```

-   The 95% CI is (-840.58, -526.25)

-   Compared to (-840.80, -526.02), we can see that the CI is a hair wider (more conservative).

## Confidence Intervals -- Sheep

-   An experiment was conducted to evaluate the effectiveness of a treatment for tapeworm in the stomachs of sheep. A random sample of 24 worm-infested lambs was randomly divided into two groups: half were injected with the drug and the other half were untreated. After 6 months, the worm counts were as follows:

|           |            |            |            |            |            |            |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Treatment | 18 <br> 13 | 43 <br> 35 | 28 <br> 38 | 50 <br> 33 | 16 <br> 6  | 32 <br> 7  |
| Control   | 40 <br> 39 | 54 <br> 23 | 26 <br> 48 | 63 <br> 58 | 21 <br> 28 | 37 <br> 39 |

## Confidence Intervals -- Sheep

```{r, echo = TRUE}
worms <- c(18, 13, 43, 35, 28, 38,
           50, 33, 16,  6, 32,  7,
           40, 39, 54, 23, 26, 48, 
           63, 58, 21, 28, 37, 39)

trt <- c(rep(1, 12), rep(0, 12))

sheep <- tibble(worms, trt)

head(sheep)
```

## Confidence Intervals -- Sheep

-   Let's now find a 99% CI for the average difference in worm counts between the treated and untreated sheep.

```{r, echo = TRUE}
t.test(worms ~ trt,
       data = sheep,
       conf.level = 0.99)
```

-   The 99% CI for $\mu_{\text{Ctrl}} - \mu_{\text{Trt}}$ is (--3.16, 29.33).

## Hypothesis Testing - Definition {.smaller}

-   Assuming $\sigma_1^2 = \sigma_2^2$,

**Hypotheses**

|          Right Tail           |           Left Tail           |           Two Tails            |
|:----------------------:|:----------------------:|:----------------------:|
| $H_0: \ \mu_1- \mu_2 \le D_0$ | $H_0: \ \mu_1 - \mu_2\ge D_0$ |  $H_0: \ \mu_1 - \mu_2 = D_0$  |
| $H_1: \ \mu_1 - \mu_2 > D_0$  | $H_1: \ \mu_1 - \mu_2 < D_0$  | $H_1: \ \mu_1 - \mu_2 \ne D_0$ |

**Test Statistic**

$$ t_0 = \frac{\left(\bar{x_1}-\bar{x_2}\right) - D_0}{s_p \sqrt{\frac{1}{n_1}+\frac{1}{n_2}}} $$

-   where

    -   $D_0$ is the hypothesized average difference between groups 1 and 2,
    -   $s_p$ is the pooled standard deviation, as defined earlier,
    -   $t_0$ has df $= n_1+n_2-2$.

## Hypothesis Testing -- Definition {.smaller}

-   Assuming $\sigma_1^2 \ne \sigma_2^2$,

**Hypotheses**

|          Right Tail           |           Left Tail           |           Two Tails            |
|:----------------------:|:----------------------:|:----------------------:|
| $H_0: \ \mu_1- \mu_2 \le D_0$ | $H_0: \ \mu_1 - \mu_2\ge D_0$ |  $H_0: \ \mu_1 - \mu_2 = D_0$  |
| $H_1: \ \mu_1 - \mu_2 > D_0$  | $H_1: \ \mu_1 - \mu_2 < D_0$  | $H_1: \ \mu_1 - \mu_2 \ne D_0$ |

**Test Statistic**

$$ t_0 = \frac{\left(\bar{x_1}-\bar{x_2}\right) - D_0}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}} $$

-   where

    -   $D_0$ is the hypothesized average difference between groups 1 and 2,
    -   the df for $t_0$'s distribution are estimated using Satterthwaite's approximation, as defined before.

## Hypothesis Testing -- R Syntax

-   We again use the [`t.test()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/t.test) function to perform the test,

```{r, echo = TRUE, eval = FALSE}
t.test([continuous variable] ~ [grouping variable],
       data = [dataset])
```

-   Like before, if we want to use exact df, we can use the `var.equal` option.

## Hypothesis Testing -- Penguins

-   Let's use the appropriate test to determine if the average body mass is different between male and female penguins. Test at the $\alpha=0.05$ level.

```{r, echo = TRUE}
t.test(body_mass_g ~ sex,
       data = penguins)
```

## Hypothesis Testing -- Penguins {.smaller}

**Hypotheses**

-   $H_0: \ \mu_{\text{F}} = \mu_{\text{M}}$
-   $H_0: \ \mu_{\text{F}} \ne \mu_{\text{M}}$

**Test Statistic**

-   $t_0 = -8.555$

***p*****-Value**

-   $p < 0.001$

**Conclusion/Interpretation**

-   Reject $H_0$ at the $\alpha=0.05$ level. There is sufficient evidence to suggest that the average body mass of penguins is different for males and females.

## Hypothesis Testing -- Sheep

-   Let's now use the appropriate hypothesis test to determine if the treatment decreased the worm count in sheep. Test at the $\alpha=0.01$ level.

```{r, echo = TRUE}
t.test(worms ~ trt,
       data = sheep,
       alternative = "greater")
```

## Hypothesis Testing -- Sheep {.smaller}

**Hypotheses**

-   $H_0: \ \mu_{\text{Trt}} \ge \mu_{\text{Ctrl}}$
-   $H_0: \ \mu_{\text{Trt}} < \mu_{\text{Ctrl}}$

**Test Statistic**

-   $t_0 = 2.271$

***p*****-Value**

-   $p = 0.017$

**Conclusion/Interpretation**

-   Fail to reject $H_0$ at the $\alpha=0.01$ level. There is not sufficient evidence to suggest that the treatment reduced the number of worms in sheep.

## Assumptions

- Note that the two-sample *t*-test assumes normality.

- We will assess this with a quantile-quantile (QQ) plot.

  - We will create a QQ plot for the data in each group.

- We want to see an approximate 45&deg; line.

```{r, echo = TRUE, eval = FALSE}
[dataset] %>% ggplot(aes(sample = [outcome])) +
  stat_qq(size=3) +
  stat_qq_line() +
  theme_minimal() +
  xlab("Theoretical") +
  ylab("Sample")
```

## Assumptions -- Penguins

```{r, echo = TRUE}
penguin_m <- penguins %>% filter(sex == "male")

penguin_m %>% ggplot(aes(sample = body_mass_g)) +
  stat_qq(size=3) +
  stat_qq_line() +
  theme_minimal() +
  xlab("Theoretical") +
  ylab("Sample")
```

## Assumptions -- Penguins

```{r, echo = TRUE}
penguin_f <- penguins %>% filter(sex == "female")

penguin_f %>% ggplot(aes(sample = body_mass_g)) +
  stat_qq(size=3) +
  stat_qq_line() +
  theme_minimal() +
  xlab("Theoretical") +
  ylab("Sample")
```

## Assumptions -- Penguins

```{r, echo = FALSE}
library(ggpubr)

m <- penguin_m %>% ggplot(aes(sample = body_mass_g)) +
  stat_qq(size=3) +
  stat_qq_line() +
  theme_minimal() +
  xlab("Theoretical") +
  ylab("Sample") +
  ggtitle("Males")

f <- penguin_f %>% ggplot(aes(sample = body_mass_g)) +
  stat_qq(size=3) +
  stat_qq_line() +
  theme_minimal() +
  xlab("Theoretical") +
  ylab("Sample") +
  ggtitle("Females")

ggarrange(m, f, ncol=2, nrow=1)
```

## Assumptions -- Sheep

```{r, echo = TRUE}
sheep_ctrl <- sheep %>% filter(trt == 0)

sheep_ctrl %>% ggplot(aes(sample = worms)) +
  stat_qq(size=3) +
  stat_qq_line() +
  theme_minimal() +
  xlab("Theoretical") +
  ylab("Sample")
```

## Assumptions -- Sheep

```{r, echo = TRUE}
sheep_trt <- sheep %>% filter(trt == 1)

sheep_trt %>% ggplot(aes(sample = worms)) +
  stat_qq(size=3) +
  stat_qq_line() +
  theme_minimal() +
  xlab("Theoretical") +
  ylab("Sample")
```

## Assumptions -- Sheep

```{r, echo = FALSE}
ctrl <- sheep_ctrl %>% ggplot(aes(sample = worms)) +
  stat_qq(size=3) +
  stat_qq_line() +
  theme_minimal() +
  xlab("Theoretical") +
  ylab("Sample") +
  ggtitle("Control")

trt <- sheep_trt %>% ggplot(aes(sample = worms)) +
  stat_qq(size=3) +
  stat_qq_line() +
  theme_minimal() +
  xlab("Theoretical") +
  ylab("Sample") +
  ggtitle("Treatment")

ggarrange(ctrl, trt, ncol=2, nrow=1)
```

