---
title: "Two-Sample Means (Independent) - Week 2, Lecture 1"
author: "Your Name Here"
format: html
self-contained: true
editor: source
---

```{r setup, echo = TRUE, warnings = FALSE, message = FALSE}
library(tidyverse)
library(gsheet)
```

**A large credit union awards quarterly bonuses to its employees. To ensure fairness across departments, data on the bonuses are regularly reviewed for equality. Consider the data in the following Google Sheet, which includes the bonus percentage for each employee in two departments.**

```{r, echo = TRUE}
data <- as_tibble(gsheet2tbl("https://docs.google.com/spreadsheets/d/1q0Ouhr6hGBDp-5oyQuYGcsWTLrsayywVHJS7o17Sm6Y/edit?usp=sharing"))
```

#### 1. Summarize the data. Provide appropriate summary statistics for the study population.

```{r, echo=TRUE}

```

#### 2. Construct the 99% confidence interval for the difference in bonuses for the two departments.

```{r, echo=TRUE}

```

#### 3. Use the appropriate hypothesis test to determine if the two departments are given different bonus amounts. Test at the &alpha;=0.01 level.

```{r, echo=TRUE}

```

#### 4. (Challenge!) Construct a graph that will help you explain your findings to management.

```{r, echo=TRUE}

```

#### 5. Write a brief paragraph explaining your findings to management. Are there any actions that should be considered as a result of your analysis?








