---
title: "Two-Sample Means (Dependent) - Week 2, Lecture 2"
author: "Your Name Here"
format: html
self-contained: true
editor: source
---

```{r setup, echo = TRUE, warnings = FALSE, message = FALSE}
library(tidyverse)
library(gsheet)
```

**A new 2 hour reading comprehension program has been developed by a local tutoring center. A pilot study of 30 students has been conducted and we have data on a reading comprehension test taken before the program and a separate reading comprehension test taken after the program. Consider the data, found in the following Google Sheet.**

```{r, echo = TRUE}
data <- as_tibble(gsheet2tbl("https://docs.google.com/spreadsheets/d/1ijxX7c6qcFAw-r_mNSxC-Ps5b5njw-WtE4RcxPTySXU/edit?usp=sharing"))
```

#### 1. Summarize the data. Provide appropriate summary statistics for the study population.

```{r, echo=TRUE}

```

#### 2. Construct the 95% confidence interval for the difference in test scores. Hint: we want to discuss this in terms of the *increase* in test scores.

```{r, echo=TRUE}

```

#### 3. Use the appropriate hypothesis test to determine if the program raises comprehension test averages by at least 5 points. Test at the &alpha;=0.05 level.

```{r, echo=TRUE}

```

#### 4. (Challenge!) Construct a graph that will help you explain your findings to the tutoring center.

```{r, echo=TRUE}

```

#### 5. Write a brief paragraph explaining your findings to the tutoring center. 










