---
title: "Lab 7: July 24"
author: "Your Name Here"
format: html
self-contained: true
execute:
  echo: true
  warning: false
  message: false
editor: source
---

**A regional airline transfers passengers from small airports to a larger regional hub airport. The airline’s data analyst was assigned to estimate the revenue (in thousands of dollars) generated by each of the 22 small airports based on two variables: the distance from each airport (in miles) to the hub and the population (in hundreds) of the cities in which each of the 22 airports is located. The data are given here: [Google Sheet](https://docs.google.com/spreadsheets/d/10ZN0jMYdVsG2ucIJFSTquVRW916Jj-JmZIGQuqSQhy0/edit?usp=sharing).**

**1. Model the revenue as a function of distance and population. Remember to state the resulting model.**

```{r}

```

**2. Use the appropriate hypothesis test to determine if distance is a significant predictor of revenue. Test at the $\alpha=0.05$ level. Remember to typeset your results.**

```{r}

```

**3. Use the appropriate hypothesis test to determine if population is a significant predictor of revenue. Test at the $\alpha=0.05$ level. Remember to typeset your results.**

```{r}

```

**4. Use the appropriate hypothesis test to determine if this is a significant regression line. Test at the $\alpha=0.05$ level.**

```{r}

```

**5. Construct a visualization for the model. At the very least, you should have a scatterplot with a regression line overlaid.**

```{r}

```

**6. How many observations are considered outliers?**

```{r}

```

**7. Create a second dataset that contains observations that are *not* classified as outliers.**

```{r}

```

**8. Using the second dataset, model the revenue as a function of distance and population. Remember to state the resulting model.**

```{r}

```

**9. Use the appropriate hypothesis test to determine if distance is a significant predictor of revenue. Test at the $\alpha=0.05$ level. Remember to typeset your results.**

```{r}

```

**10. Use the appropriate hypothesis test to determine if population is a significant predictor of revenue. Test at the $\alpha=0.05$ level. Remember to typeset your results.**

```{r}

```

**11. Use the appropriate hypothesis test to determine if this is a significant regression line. Test at the $\alpha=0.05$ level.**

```{r}

```

**12. Construct a visualization for the model. At the very least, you should have a scatterplot with a regression line overlaid.**

```{r}

```

**13. Why do you think I chose this as your lab activity? What are you taking away from the analyses you have done?**








