---
title: "ANOVA Assumptions and the Kruskal-Wallis - Week 5, Lecture 2"
author: "Your Name Here"
format: html
self-contained: true
editor: source
---

**Recall the agricultural data from the ANOVA activity available in the following [Google Sheet](https://docs.google.com/spreadsheets/d/1CFyuuWy-4YmVU3wkYuC8Uf3xW3qkpCnpF468NjBAhbY/edit#gid=0).**

#### 1. State the ANOVA assumptions mathematically.

#### 2. Explain how to check the ANOVA assumptions.

```{r, echo=TRUE}

```

#### 3. Assess the ANOVA assumptions for the agricultural data, graphically. In addition to providing the relevant graphs, you must provide an assessment for each assumption.

```{r, echo=TRUE}

```

#### 4. Use the appropriate hypothesis test to assess the variance assumption. Test at the $\alpha=0.05$ level.

```{r, echo=TRUE}

```

#### 5. Provide an overall assessment -- do we meet the assumptions to use ANOVA?

#### 6. Perform the Kruskal-Wallis to determine if there is a difference between the treatment groups. Test at the $\alpha=0.05$ level.

```{r, echo=TRUE}

```

#### 7. Perform the Kruskal-Wallis post-hoc test to determine what differences exist between groups. Test at the $\alpha=0.05$ level.

#### 8. Challenge! Construct a graph (or multiple graphs) to aid in explaining this to the agricultural researcher.

#### 9. Write a summary paragraph, appropriate for the agricultural researcher (i.e., not a statistician), using the results that are **appropriate for the data**. Hint: you will summarize either your ANOVA findings from the previous activity *or* your Kruskal-Wallis findings from this activity.


