---
title: "Logistic Regression - Week 12, Day 1"
author: "Your Name Here"
format: html
self-contained: true
editor: source
---

**Recall the Titanic data from lecture. Use it to answer the following questions.**

**1. Model survival as a function of age (*Age*), the number of siblings and/or spouses (*SibSp*), the number of parents or children (*Parch*), the class of ticket (*Pclass*), and amount paid for ticket (*Fare*).** 

```{r, echo = TRUE}

```

**1a. Report the model.**

**1b. Which, if any, are significant predictors of survival? The full hypothesis test is not needed, but you must provide statistical justification.**

```{r, echo = TRUE}

```

**1c. Find the odds ratios and corresponding 95% confidence intervals.**

```{r, echo = TRUE}

```

**1d. Provide interpretations for the odds ratios for each of the predictors in the model.**

**2. Challenge! Model survival as a function of age (*Age*), the class of ticket (*Pclass*), and sex of passenger (*Sex*).** 

```{r, echo = TRUE}

```

**2a. Report the model.**

**2b. Which, if any, are significant predictors of survival? The full hypothesis test is not needed, but you must provide statistical justification.**

```{r, echo = TRUE}

```

**2c. Find the odds ratios and corresponding 95% confidence intervals.**

```{r, echo = TRUE}

```

**2d. Provide interpretations for the odds ratios for each of the predictors in the model.**
