---
title: "Logistic Regression - Week 13, Day 2"
author: "Your Name Here"
format: html
self-contained: true
editor: source
---

**1. Challenge! Research "good" values of the metrics we discussed in lecture. How can we determine how well the model is performing using these metrics?**

**2. Recall question 1 from the last activity.**

**2a. Restate the model from 1a here.** 

```{r, echo = TRUE}

```

**2b. Find the confusion matrix for this model.**

```{r, echo = TRUE}

```

**2c. Find the sensitivity of the model.**

```{r, echo = TRUE}

```

**2d. Find the specificity of the model.**

```{r, echo = TRUE}

```

**2e. Find the positive predictive value of the model.**

```{r, echo = TRUE}

```

**2f. Find the negative predictive value of the model.**

```{r, echo = TRUE}

```

**2g. Find the false discovery rate of the model.**

```{r, echo = TRUE}

```

**3. Compare and contrasts the results from the models constructed in  lecture and the last activity. Which model performs better, based on your answer in question 1?**

**4. Challenge! Construct a graph to support your answer in question 3.**

```{r, echo = TRUE}

```




