---
title: "One Sample Means"
subtitle: "STA5176: Statistical Modeling"
author: "Dr. Seals"
format: 
  revealjs:
    theme: dark2
    self-contained: false
    slide-number: false
    footer: "[STA5176 - Statistical Modeling](https://samanthaseals.github.io/STA5176)"
    width: 1600
    height: 900
    chalkboard: true
    df-print: paged
    html-math-method: katex
editor: source
---

```{r setup_environment, include = FALSE}
library("tidyverse") 
```

## Introduction

-   In the last lecture, we reviewed summary statistics and the concepts of statistical inference (confidence intervals and hypothesis testing).

-   Today we will cover estimation and inference on one mean using the one-sample *t*-test.

-   We will only discuss the *t*-test approach, but note that under certain conditions, we could use a *z*-test.

## Palmer Penguins

<div style = "font-size: .75em; margin-top: 2em">

```{r}
penguins <- palmerpenguins::penguins
set.seed(1)
penguins |>
 slice_sample(prop = 1)
```

## Confidence Interval -- Definition

-   The CI for $\mu$ is as follows:

$$ \bar{x} \pm t_{1-\alpha/2, \ n-1} \frac{s}{\sqrt{n}} $$

-   We will use the [`t.test()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/t.test) function to find the CI,

```{r, echo = TRUE, eval = FALSE}
t.test([dataset]$[variable], conf.level = [confidence level])
```

## Confidence Interval -- Penguins

-   Find the 95% CI for the body mass of penguins.

```{r, echo = TRUE}
t.test(penguins$body_mass_g, conf.level = 0.95)
```

-   The 95% CI for $\mu$ is (4116.46 grams, 4287.05 grams)

## Confidence Interval -- Pilots

-   An airline wants to evaluate the depth perception of its pilots over the age of 50. A random sample of 14 airline pilots over the age of 50 are asked to judge the distance between two markers placed 20 feet apart at the opposite end of the laboratory. The sample data listed here are the pilots' error (recorded in feet) in judging the distance. Use the sample data to place a 99% confidence interval on the average error in depth perception for the company's pilots over the age of 50.

|     |     |     |     |     |     |     |
|-----|-----|-----|-----|-----|-----|-----|
| 2.7 | 2.4 | 1.9 | 2.6 | 2.4 | 1.9 | 2.3 |
| 2.2 | 2.5 | 2.3 | 1.8 | 2.5 | 2.0 | 2.2 |

## Confidence Interval -- Pilots

```{r, echo = TRUE}
error <- c(2.7, 2.4, 1.9, 2.6, 2.4, 1.9, 2.3,
           2.2, 2.5, 2.3, 1.8, 2.5, 2.0, 2.2)

t.test(error, conf.level = 0.99)
```

-   The 99% CI for $\mu$ is (2.04, 2.49).

## Confidence Interval -- Salmonella

-   A massive Salmonella outbreak occurred due to ice cream. Epidemiologists sampled nine production runs from the company that had produced the ice cream to determine the level of Salmonella in the ice cream. Construct the appropriate 90% CI and determine whether the average level of Salmonella in the ice cream is greater than 0.3, a level that is considered to be very dangerous.

|       |       |       |
|-------|-------|-------|
| 0.593 | 0.142 | 0.329 |
| 0.691 | 0.231 | 0.793 |
| 0.519 | 0.392 | 0.418 |

## Confidence Interval -- Salmonella

```{r, echo = TRUE}
salmonella <- c(0.593, 0.142, 0.329, 
                0.691, 0.231, 0.793, 
                0.519, 0.392, 0.418)

t.test(salmonella, conf.level = 0.90)
```

-   The 95% CI for $\mu$ is (0.325, 0.588).
-   The CI is entirely above 0.3, thus, we can conclude that the level of Salmonella in the ice cream is larger than 0.3.

## Hypothesis Testing - Definition {.smaller}

**Hypotheses**

|       Right Tail       |       Left Tail        |       Two Tails        |
|:----------------------:|:----------------------:|:----------------------:|
| $H_0: \ \mu \le \mu_0$ | $H_0: \ \mu \ge \mu_0$ |  $H_0: \ \mu = \mu_0$  |
|  $H_1: \ \mu > \mu_0$  |  $H_1: \ \mu < \mu_0$  | $H_1: \ \mu \ne \mu_0$ |

**Test Statistic**

$$ t_0 = \frac{\bar{x}-\mu_0}{s / \sqrt{n}} $$

-   We again use the [`t.test()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/t.test) function to perform the test,

```{r, echo = TRUE, eval = FALSE}
t.test([dataset]$[variable], 
       mu = [hypothesized value], 
       alternative = [alternative])
```

## Hypothesis Testing -- Penguins

-   Recall the penguin data. Determine if the average body mass of penguins is different from 4400 grams. Test at the $\alpha=0.05$ level.

```{r, echo = TRUE}
t.test(penguins$body_mass_g, mu=4400, alternative = "two")
```

## Hypothesis Testing -- Penguins 

**Hypotheses**

-   $H_0: \ \mu = 4400$ grams
-   $H_1: \ \mu \ne 4400$ grams

**Test Statistic**

-   $t_0 = -4.572$

***p*****-Value**

-   $p < 0.001$

**Conclusion/Interpretation**

-   Reject $H_0$ at the $\alpha=0.05$ level. There is sufficient evidence to suggest that the average body mass of penguins is different from 4400 grams.

## Hypothesis Testing -- Salmonella

-   Recall the Salmonella data. Determine if the Salmonella levels are greater than 0.3. Test at the $\alpha=0.10$ level.

```{r, echo = TRUE}
t.test(salmonella, mu=0.3, alternative = "greater")
```

## Hypothesis Testing -- Salmonella

**Hypotheses**

-   $H_0: \ \mu \le 0.3$
-   $H_1: \ \mu > 0.3$

**Test Statistic**

-   $t_0 = 2.205$

***p*****-Value**

-   $p = 0.039$

**Conclusion/Interpretation**

-   Reject $H_0$ at the $\alpha=0.10$ level. There is sufficient evidence to suggest that the average Salmonella level is greater than 0.3.
