---
title: "Inference on Variances"
subtitle: "STA5176: Statistical Modeling"
author: "Dr. Seals"
format: 
  revealjs:
    theme: dark2
    self-contained: false
    slide-number: false
    footer: "[STA5176 - Statistical Modeling](https://samanthaseals.github.io/STA5176)"
    width: 1600
    height: 900
    chalkboard: true
    df-print: paged
    html-math-method: katex
editor: source
---

## Introduction

```{r, echo = FALSE}
library(tidyverse)
library(EnvStats)
library(car)
penguins <- palmerpenguins::penguins
```

-   In the last lecture, we covered estimation and inference on central parameters ($\mu$ and M).

-   Today we will cover estimation and inference on $\sigma^2$

-   We will learn the tests for one, two, and more than two variances

    -   Typically, these are used to check assumptions of either the $t$-test (two variances) or ANOVA (more than two variances).
    
## Hypothesis Testing -- One $\sigma^2$

**Hypotheses**

|        Right Tail         |         Left Tail         |         Two Tails          |
|:----------------------:|:----------------------:|:----------------------:|
| $H_0: \ \sigma^2 \le \sigma^2_0$ | $H_0: \sigma^2 \ge \sigma^2_0$ |  $H_0: \ \sigma^2 = \sigma^2_0$  |
| $H_1: \ \sigma^2 > \sigma^2_0$  | $H_1: \ \sigma^2 < \sigma^2_0$  | $H_1: \ \sigma^2 \ne \sigma^2_0$ |

**Test Statistic**

$$ \chi^2_0 = \frac{(n-1)s^2}{\sigma^2_0} $$

## Hypothesis Testing -- One $\sigma^2$

  -   We will  use the [`varTest()`](https://www.rdocumentation.org/packages/EnvStats/versions/2.3.1/topics/varTest) function from the [`EnvStats`](https://www.rdocumentation.org/packages/EnvStats/versions/2.7.0/topics/EnvStats-package) package to perform the test,

```{r, echo = TRUE, eval = FALSE}
varTest([variable], 
        alternative = "[alternative]", 
        conf.level = [level],
        sigma.squared = [hypothesized value])
```

- The code will not run without the `conf.level` stated.

- Note that if you are constructing a two-tailed test, you can find the corresponding CI in the same call of `varTest`.

- If you are constructing a one-tailed test, you will see that the CI presented is a one-sided CI, while we are interested in two-sided CIs.

## Hypothesis Testing -- One $\sigma^2$ -- Penguins

-   Recall the penguin data. Let's use the appropriate test to determine if the standard deviation of bill length for penguins (`r round(sd(penguins$bill_length_mm, na.rm = TRUE),2)` mm) is greater than 5 mm. Test at the $\alpha=0.01$ level.

```{r, warning = FALSE, echo = TRUE}
varTest(penguins$bill_length_mm, 
        alternative = "greater", 
        conf.level = 0.95, 
        sigma.squared = 25)
```

## Hypothesis Testing -- One $\sigma^2$ -- Penguins

**Hypotheses**

-   $H_0: \ \sigma \le 5$ OR $\sigma^2 \le 25$
-   $H_1: \ \sigma > 5$ OR $\sigma^2 > 25$

**Test Statistic**

-   $\chi^2_0 = 406.57$

***p*****-Value**

-   $p = 0.008$

**Conclusion/Interpretation**

-   Reject $H_0$ at the $\alpha=0.01$ level. There is sufficient evidence to suggest that the standard deviation of bill length is larger than 5 mm.

## Confidence Intervals -- One $\sigma^2$

  - For variance:
$$ \left( \frac{(n-1)s^2}{\chi^2_{1-\alpha/2, n-1}}, \frac{(n-1)s^2}{\chi^2_{\alpha/2, n-1}} \right) $$

  - For standard deviation:
$$ \left( \sqrt{\frac{(n-1)s^2}{\chi^2_{1-\alpha/2, n-1}}}, \sqrt{\frac{(n-1)s^2}{\chi^2_{\alpha/2, n-1}}} \right) $$

## Confidence Intervals -- One $\sigma^2$

  -   We will use the [`varTest()`](https://www.rdocumentation.org/packages/EnvStats/versions/2.3.1/topics/varTest) function to find the CI,

```{r, echo = TRUE, eval = FALSE}
varTest([variable], 
        conf.level = [level])
```

- Note that we now have removed the `alternative` and `sigma.squared` options.
  
    - The default `alternative` is two-sided.
    - The default `sigma.squared` is 1.

## Confidence Intervals -- One $\sigma^2$ -- Penguins

-   Recall the penguin data. Find the 99% CI for the standard deviation, $\sigma$.

```{r, warning = FALSE, echo = TRUE}
varTest(penguins$bill_length_mm, 
        conf.level = 0.99)
```

- The 99% CI for $\sigma^2$ is (24.67, 36.63)

- However, we were asked for the 99% CI for $\sigma$, so we must take the square root.

    - Thus, the 99% CI for $\sigma$ is (`r round(sqrt(24.67),2)`, `r round(sqrt(36.63),2)`).


## Hypothesis Testing -- Two $\sigma^2$

  - This is known as the folded $F$ test.

**Hypotheses**

|        Right Tail         |         Left Tail         |         Two Tails          |
|:----------------------:|:----------------------:|:----------------------:|
| $H_0: \ \sigma_1^2 \le \sigma^2_2$ | $H_0: \sigma_1^2 \ge \sigma^2_2$ |  $H_0: \ \sigma_1^2 = \sigma^2_2$  |
| $H_1: \ \sigma_1^2 > \sigma^2_2$  | $H_1: \ \sigma_1^2 < \sigma_2^2$  | $H_1: \ \sigma_1^2 \ne \sigma_2^2$ |

**Test Statistic**

|        Right Tail         |         Left Tail         |         Two Tails          |
|:----------------------:|:----------------------:|:----------------------:|
| $F_0 = \frac{s_1^2}{s_2^2}$ | $F_0 = \frac{s_2^2}{s_1^2}$ |  $F_0 = \frac{s_1^2}{s_2^2}$  |

## Hypothesis Testing -- Two $\sigma^2$

- We will use the [`var.test()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/var.test) function from the [`stats`](https://www.rdocumentation.org/packages/stats/versions/3.6.2) package.

```{r, echo = TRUE, eval = FALSE}
var.test([var 1], [var 2], 
         alternative = [alternative], 
         conf.level = [level], 
         ratio = [ratio (normally 1)])
```

- We do not have to specify the `conf.level` if only interested in the hypothesis test.

- Remember that this is typically used to verify the variance assumption of the two-sample $t$-test.

## Hypothesis Testing -- Two $\sigma^2$ -- Penguins

- Recall the penguins data. Determine if the variance of the bill length is the same between male and female penguins. Test at the $\alpha=0.05$ level.

```{r, echo = TRUE}
f <- penguins %>% filter(sex == "female")
m <- penguins %>% filter(sex == "male")
var.test(f$bill_length_mm, m$bill_length_mm,
         alternative = "two",
         ratio = 1)
```

## Hypothesis Testing -- Two $\sigma^2$ -- Penguins

**Hypotheses**

-   $H_0: \ \sigma^2_{\text{M}} = \sigma^2_{\text{F}}$
-   $H_1: \ \sigma^2_{\text{M}} \ne \sigma^2_{\text{F}}$

**Test Statistic**

-   $F_0 = 0.83$

***p*****-Value**

-   $p = 0.247$

**Conclusion/Interpretation**

-   Fail to reject $H_0$ at the $\alpha=0.05$ level. There is not sufficient evidence to suggest that the variances are different.

## Confidence Intervals -- Two $\sigma^2$

- For the ratio of two variances:
$$ \left( \frac{s_1^2}{s_2^2} F_{1-\alpha/2, n_2-1, n_1-1}, \frac{s_1^2}{s_2^2} F_{\alpha/2, n_2-1, n_1-1} \right)$$

## Confidence Intervals -- Two $\sigma^2$

- We again use the [`var.test()`](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/var.test) function from the [`stats`](https://www.rdocumentation.org/packages/stats/versions/3.6.2) package.

```{r, echo = TRUE, eval = FALSE}
var.test([var 1], [var 2], 
         conf.level = [level])
```

## Confidence Intervals -- Two $\sigma^2$

- Let's find the 95% CI for the ratio of variances between male and female penguins.

```{r, echo = TRUE}
var.test(f$bill_length_mm, m$bill_length_mm,
         conf.level = 0.95)
```

- Thus, the 95% CI for the ratio of variances is (0.61, 1.13).

## Hypothesis Testing --  $\ge$ Three $\sigma^2$

- The Brown-Forsythe-Levine test is used to assess assumptions in ANOVA (next chapter).

    - In fact, it is transforming the data, then performing ANOVA on the transformed data.

**Hypotheses**

  - $H_0: \ \sigma^2_1 = ... = \sigma^2_k$
  - $H_1:$ at least one $\sigma^2_i$ is different
  

## Hypothesis Testing -- $\ge$ Three $\sigma^2$

**Test Statistic**
$$ F_0 = \frac{\sum_{i=1}^k n_i (\bar{z}_i - \bar{z})^2/(k-1)}{\sum_{i=1}^k \sum_{j=1}^{n_j}(z_{ij}-\bar{z}_i)^2/(n-k) }, $$

where

  - $k$ is the number of groups,
  - $n_i$ is the sample size of group i,
  - $n = \sum_{i=1}^k n_i$, and
  - $z_{ij} = |y_{ij} - \text{median}(y_i)|$
  
## Hypothesis Testing -- $\ge$ Three $\sigma^2$

- We will use the [`leveneTest()`](https://www.rdocumentation.org/packages/car/versions/3.1-0/topics/leveneTest) function from the [`car`](https://www.rdocumentation.org/packages/car/versions/3.1-0) package.

```{r, echo = TRUE, eval = FALSE}
leveneTest([continuous var] ~ [grouping var], 
           data = [dataset], 
           center = median)
```
  
## Hypothesis Testing -- $\ge$ Three $\sigma^2$ -- Penguins

- Let's now determine if the variance of bill length is different between the three species of penguins.

```{r, echo = TRUE}
leveneTest(bill_length_mm ~ species, data = penguins, center = median)
```

## Hypothesis Testing -- $\ge$ Three $\sigma^2$ -- Penguins

**Hypotheses**

-   $H_0: \ \sigma^2_{\text{Adelie}} = \sigma^2_{\text{Chinstrap}} = \sigma^2_{\text{Gentoo}}$
-   $H_1:$ at least one $\sigma^2_i$ is different

**Test Statistic**

-   $F_0 = 2.24$

***p*****-Value**

-   $p = 0.108$

**Conclusion/Interpretation**

-   Fail to reject $H_0$ at the $\alpha=0.05$ level. There is not sufficient evidence to suggest that the variances are different.




