---
title: "Multiple Regression: Estimation and Inference - Week 9, Lecture 2"
author: "Dr. Seals"
format: html
self-contained: true
editor: source
---

**A regional airline transfers passengers from small airports to a larger regional hub airport. The airline's data analyst was assigned to estimate the revenue (in thousands of dollars) generated by each of the 22 small airports based on two variables: the distance from each airport (in miles) to the hub and the population (in hundreds) of the cities in which each of the 22 airports is located. The data are given here: [Google Sheet](https://docs.google.com/spreadsheets/d/10ZN0jMYdVsG2ucIJFSTquVRW916Jj-JmZIGQuqSQhy0/edit?usp=sharing).**

```{r setup, echo = TRUE, warnings = FALSE, message = FALSE}
library(tidyverse)
library(gsheet)
data <- as_tibble(gsheet2tbl("https://docs.google.com/spreadsheets/d/10ZN0jMYdVsG2ucIJFSTquVRW916Jj-JmZIGQuqSQhy0/edit?usp=sharing"))
```

**1. Summarize the data using appropriate summary statistics.**

**2. Construct the correlation matrix for revenue, distance, and population.**

```{r, echo = TRUE}
cor(data[2:4])
```

**3. Are any correlations in Q2 suspiciously high?**

Only population and revenue have a high correlation, however, revenue is the outcome of interest, so it is okay that the correlation is high.

**4.  Model the revenue as a function of distance and population.**

```{r, echo = TRUE}
m1 <- lm(Revenue ~ Distance + Population, data = data)
coefficients(m1)
```

$$\hat{\text{revenue}} = 144.38 +  0.17 \text{ distance} + 1.09 \text{ population}$$

**5.  Provide appropriate interpretations for the model coefficients.**

-   Without considering distance and population, the average revenue is \$144,388.

-   For a 1 mile increase in distance, revenue increases by \$169.

-   For a 100 person increase in the population, revenue increases by \$1090.

    -   Note that it is a 100 person increase because the problem set up tells us that the population is in terms of 100 people.

**6.  Find the 95% confidence interval for all $\beta_i$.**

```{r, echo = TRUE}
confint(m1)
```

-   The 95% CI for $\beta_{\text{distance}}$ is (-0.149, 0.487)

-   The 95% CI for $\beta_{\text{population}}$ is (0.671, 1.518)

**7.  Use the appropriate hypothesis test to determine if distance is a significant predictor of revenue. Test at the $\alpha=0.05$ level.**

```{r, echo = TRUE}
summary(m1)
```

**Hypotheses**

-   $H_0: \ \beta_{\text{distance}} = 0$
-   $H_1: \ \beta_{\text{distance}} \ne 0$

**Test Statistic and *p*-Value**

-   $t_0 = 1.114$
-   $p = 0.279$

**Rejection Region**

-   Reject $H_0$ if $p<\alpha$; $\alpha=0.05$

**Conclusion/Interpretation**

-   Fail to reject $H_0$ at the $\alpha=0.05$ level. There is not sufficient evidence to suggest that distance is a significant predictor of revenue.

**8.  Use the appropriate hypothesis test to determine if population is a significant predictor of revenue. Test at the $\alpha=0.05$ level.**

```{r, echo = TRUE}
summary(m1)
```

**Hypotheses**

-   $H_0: \ \beta_{\text{distance}} = 0$
-   $H_1: \ \beta_{\text{distance}} \ne 0$

**Test Statistic and *p*-Value**

-   $t_0 = 1.114$
-   $p = 0.279$

**Rejection Region**

-   Reject $H_0$ if $p<\alpha$; $\alpha=0.05$

**Conclusion/Interpretation**

-   Fail to reject $H_0$ at the $\alpha=0.05$ level. There is not sufficient evidence to suggest that distance is a significant predictor of revenue.

**9.  Use the appropriate hypothesis test to determine if this is a significant regression line. Test at the $\alpha=0.05$ level.**

```{r, echo = TRUE}
summary(m1)
```

**Hypotheses**

-   $H_0: \ \beta_{\text{distance}} =  \beta_{\text{population}} = 0$
-   $H_1 \ $ at least one $\beta_i \ne 0$

**Test Statistic and *p*-Value**

-   $F_0 = 14.65$
-   $p < 0.001$

**Rejection Region**

-   Reject $H_0$ if $p<\alpha$; $\alpha=0.05$

**Conclusion/Interpretation**

-   Reject $H_0$ at the $\alpha=0.05$ level. There is sufficient evidence to suggest that at least one slope is non-zero.

**10. Find the $R^2_{\text{adj}}$ value and provide an interpretation.**

```{r, echo = TRUE}
summary(m1)
```

$R^2_{\text{adj}} = 0.5653$; ~ 57% of the variance in revenue is explained by distance and population.

**11. Construct a visualization for the model. At the very least, you should have a scatterplot with a regression line overlaid.**

