---
title: "Goodness of Fit - Week 12, Day 1"
author: "Dr. Seals"
format: html
self-contained: true
editor: source
---

**1. The quality control department of a motorcycle company classifies new motorcycles according to the number of defective components per motorcycle at an initial inspection. An improvement to the production process has been implemented, and, hopefully, there will be a change from the historical defective distribution: $\pi_1 = 0.80,$ $\pi_2 = 0.10,$ $\pi_3 = .05,$ $\pi_4 = .03,$ and $\pi_5 = 0.02$. A random sample of 300 motorcycles produced under the new system is classified as follows:**

| No. Defective | No. Motorcycles |
|---|---|
| 0 | 238 |
| 1 | 32 |
| 2 | 12 |
| 3 | 13 |
| 4+ | 5 |

**At the \alpha = 0.05$ level, does there appear to be a change in the historical proportions of defectives?**

```{r, echo = TRUE}
counts <- c(238, 32, 12, 13, 5)
probs <- c(0.80, 0.10, 0.05, 0.03, 0.02)
chisq.test(counts, p = probs)
```

**Hypotheses**

- $H_0: \ $  the number of defective components per motorcycle is not different from the historical distribution
- $H_1: \ $ the number of defective components per motorcycle is different from the historical distribution

**Test Statistic and *p*-Value**

- $\chi^2_0 = 2.69$
- $p = 0.610$

**Rejection Region**

- Reject $H_0$ if $p < \alpha$; $\alpha=0.05$

**Conclusion/Interpretation**

- Fail to reject $H_0$ at the $\alpha=0.05$ level. There is not sufficient evidence to suggest that the number of defective components per motorcycle is different from the historical distribution.

**2. The data in the following table document the starting positions of the winning horses in 144 races. The starting position listed as 1 is the position of the horse in the starting gate closest to the inside rail of the track, and position 8 is farthest from the rail. Racing officials contend that starting position has no effect on the chance of winning the race.**

| Starting Position | No. Winners |
|---|---|
| 1 | 29 |
| 2 | 19 |
| 3 | 18 |
| 4 | 25 |
| 5 | 17 |
| 6 | 10 |
| 7 | 15 |
| 8 | 11 |

**Do the data support the racing officials' contention?**

```{r, echo = TRUE}
counts <- c(29, 19, 18, 25, 17, 10, 15, 11)
probs <- c(rep(1/8, 8))
chisq.test(counts, p = probs)
```

**Hypotheses**

- $H_0: \ $  starting position has no effect on the chance of winning (i.e., the data comes from a uniform distribution)
- $H_1: \ $ starting position has an effect on the chance of winning (i.e., the data does not come from a uniform distribution)

**Test Statistic and *p*-Value**

- $\chi^2_0 = 16.33$
- $p = 0.022$

**Rejection Region**

- Reject $H_0$ if $p < \alpha$; $\alpha=0.05$

**Conclusion/Interpretation**

- Reject $H_0$ at the $\alpha=0.05$ level. There is sufficient evidence to suggest that the starting position has an effect on winning.

**3. Challenge! A study examining bomb hits in South London during World War II is documented in the following table. The bomb hits were recorded in the 576 grids in a map of a region in South London. The study contended that certain areas were less likely to be hit with a bomb because of certain geographical features. If the bomb hits were purely random, a Poisson model would produce the number of hits per grid.**

| No. Bomb Hits | No. Grids |
|---|---|
| 0 | 229 |
| 1 | 211 |
| 2 | 93 |
| 3 | 35 |
| 4 | 7 |
| 5 | 0 |
| 6 | 0 |
| 7 | 1 |

**Does the distribution of bomb hits appear to be random across this region of South London?**

We will follow the example that starts on page 507 in the textbook.

First, we must estimate $\mu$,

```{r, echo = TRUE}
x_bar <- (229*0+211*1+93*2+35*3+7*4+0*5+0*6+1*7)/(229+211+93+35+7+0+0+1)
```

This gives us an estimated mean of `r round(x_bar, 4)`.

Now, we need to find the Poisson probabilities under $\mu=\lambda=`r round(x_bar, 4)`$,

```{r, echo = TRUE, warning = FALSE, message = FALSE}
library(tidyverse)
x <- seq(0, 6, 1) # exclude 7 for "tail end" purposes
exp_p <- dpois(x, x_bar)
observed <- c(229, 211, 93, 35, 7, 0, 0)
data <- tibble(x, observed, exp_p)

# this part is necessary because otherwise, we do not have 
# probabilities that sum to 1
x_7 = 7
observed_7 = 1
exp_p_7 = 1 - ppois(6, x_bar)
last_row <- tibble(x_7, observed_7, exp_p_7) %>%
  rename(x = x_7, 
         observed = observed_7,
         exp_p = exp_p_7)

data <- rbind(data, last_row)
```

Now that we have the Poisson probabilities, we can use the goodness of fit.

```{r, echo = TRUE}
chisq.test(data$observed, p = data$exp_p)
```

**Hypotheses**

- $H_0: \ $  the bomb hits are purely random (i.e., follow a Poisson distribution)
- $H_1: \ $ the bomb hits are not purely random (i.e., do not follow a Poisson distribution)

**Test Statistic and *p*-Value**

- $\chi^2_0 = 32.67$
- $p < 0.001$

**Rejection Region**

- Reject $H_0$ if $p < \alpha$; $\alpha=0.05$

**Conclusion/Interpretation**

- Reject $H_0$ at the $\alpha=0.05$ level. There is sufficient evidence to suggest that the bomb hits were not purely random.